Updates a Procrustes and a PCA

### 2025-05-02 Analysis/17-update_procrustes_and_PCA.Rmd
Working-ish?

## Project them! (*-1 is to flip it around)
updated_pca <- update.gpa.pca(whole_gpa, new_individuals)*-1

## From Morpho:
eigenPCA <- function(data,tol=.Machine$double.eps) {
    center1 <- colMeans(data,2,mean)
    datascale <- scale(data,2,scale=FALSE)
    eigData <- eigen(cov(datascale),symmetric=TRUE)
    eigData$values <- eigData$values[which(eigData$values > tol)] 
    sdev <- sqrt(eigData$values)
    rotation <- eigData$vectors[,1:length(sdev)]
    x <- t(t(rotation)%*%t(datascale))
    return(list(x=x,rotation=rotation,sdev=sdev,center=center1))
}

# No pipeline
# Get the CPA
original_data <- whole_gpa
new_landmarks <- new_individuals
original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
original_pca <- prcompfast(original_data$tan*-1) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
original_pca <- eigenPCA(original_data$tan)

    center1 <- colMeans(data,2,mean)
    datascale <- scale(data,2,scale=FALSE)
    eigData <- eigen(cov(datascale),symmetric=TRUE)

landvR::array.to(original_data$rotated, to = "matrix")

## Project the new specimens on the original GPA
new_data <- Morpho::align2procSym(original_data, new_landmarks)
## Project the new GPAed specimens onto the PCA
new_pca <- stats::predict(original_pca, newdata = array.to(new_data, to = "matrix"))
## Combine the new and old matrices for the output PCA
updated_pca <- rbind(original_pca$x, new_pca) *-1

## Some big shift!
plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
## Plot the chaffinches to see where they lay:
fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))
```

>>>>

## Project them! (*-1 is to flip it around)
# updated_pca <- update.gpa.pca(whole_gpa, new_individuals)*-1

# No pipeline
# Get the CPA
original_data <- whole_gpa
new_landmarks <- new_individuals
# original_pca <- prcomp(landvR::array.to(original_data$rotated, to = "matrix")) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
original_pca <- prcompfast(original_data$tan*-1) # SOLUTION IS SOMEWHERE BETWEEN THIS AND THE PRCOMPFAST
#original_pca <- eigenPCA(landvR::array.to(original_data$rotated, to = "matrix"))

    # center1 <- colMeans(data,2,mean)
    # datascale <- scale(data,2,scale=FALSE)
    # eigData <- eigen(cov(datascale),symmetric=TRUE)

## Project the new specimens on the original GPA
new_data <- Morpho::align2procSym(original_data, new_landmarks)
## Project the new GPAed specimens onto the PCA

## New data needs to be projected on the tangent space because of symmetry.
# new_procrustes <- array.to(new_data, to = "matrix")

new_proj_symtan <- sweep(new_data, 1:2, whole_gpa$mshape)
new_proc_tan <- vecx(new_proj_symtan)
# if (sizeshape) {
#     meanlogCS <- mean(log(CS))
#     CSlog <- log(CS) - meanlogCS
#     tan <- cbind(CSlog,tan)
# }

new_pca <- stats::predict(original_pca, newdata = new_proc_tan)#array.to(new_proc_tan, to = "matrix"))
## Combine the new and old matrices for the output PCA
updated_pca <- rbind(original_pca$x*-1, new_pca)

## Some big shift!
plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
## Plot the chaffinches to see where they lay:
fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))
```
